---
title: "Assignment Classes 03-07"
author: "GH"
date: "2024-10-22"
output:
  word_document:
    reference_docx: MB2-Rstudio-template.docx
    toc: true
  html_document:
    df_print: paged
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Authors of this assignment
Please state name(s) and Matrikelnummer of all (up to 3) students who contributed to this assignment!

1. Michael Adam, 01507402
2. Aaron Paul Osburg, 12010313


## Exercise 1

Develop a strategy to improve the linear regression model for this data set. What could be done to achieve a model fit that has less issues with residuals? First, develop a plan with several analysis strategies and summarize these ideas in written form. Second, carry out the analyses and check the resulting model(s) again. Third, describe your new model(s) appropriately. Fourth, write a summary discussing successes/failures of your strategies.

Written description:
The residuals are not normally distributed as the QQ-Plot of assignment 1 showed. This could be due to a wrong or incomplete choice of predictors or a non-normal distribution of the dependent variable. If the distribution of the dependent variable is
not normal, one could try to apply a transformation such as a Box-Cox, a square root or a polynomial transformation.
Otherwise, or if the problems remain after applying such a transformation, additional or different predictors (such as interaction and higher order terms, splines, or fractional polynomials) could be used instead. 

Potentially better predictors could be found by using AIC-based model selection algorithms (starting from a model with higher-order terms such as a fully-quadratic model) or inspecting partial residual plots of the
original predictors to deduce a suitable nonlinear transformation.  

Perform analysis:
```{r}
library(mfp)
library(splines)
library(MASS)
url_diab <- "https://hbiostat.org/data/repo/diabetes.csv"
diabetes <- read.csv(file=url_diab)
diabetes$BMI <-  diabetes$weight / diabetes$height^2 * 704.5  

hist(diabetes$glyhb)
```
Clearly the distribution of glyhb strongly deviates from a normal distribution and is highly assymetric with a long tail on the right side. Therefore, we now determine a Box-Cox transformation starting from the model from assignment 1.

```{r}
model <- lm(glyhb ~ chol + hdl + age + gender + waist + bp.1s + BMI, data=diabetes)
b <- boxcox(model)
lambda <- b$x[which.max(b$y)]
lambda

hist((diabetes$glyhb^lambda - 1)/lambda)
```
The distribution of the dependent vairable looks much more symmetric and Gaussian-like now, even though it still does not perfectly look like a Gaussian. 

```{r}
model <- lm((glyhb^lambda - 1)/lambda ~ chol + hdl + age + gender + waist + bp.1s + BMI, data=diabetes)
summary(model)
plot(model)
```

The QQ-Plot shows that the are approximately normally distributed now. Deviations can only be seen in the tails.
The scale-location plot shows that the tendency of the residuals to get larger for larger fitted values decreased compared to the original model from assignment 1.

To further improve the model, we try to improve our choice of predictors by applying an AIC-based model selection
starting from a fully-quadratic model (except for the quadratic term of the binary predictor gender).

```{r}
model <- lm((glyhb^lambda - 1)/lambda ~ (chol + hdl + age + gender + waist + bp.1s + BMI)^2 + I(chol^2) + I(hdl^2) + I(age^2) + I(waist^2) + I(bp.1s^2) + I(BMI^2), 
data=na.omit(diabetes))
model <- step(model)

#model with the terms resulting from the reduction
model <- lm((glyhb^lambda - 1)/lambda ~ chol + hdl + age + gender + waist + bp.1s + BMI + I(BMI^2) + chol:hdl + chol:BMI + age:bp.1s + age:gender + waist:bp.1s, data=diabetes)

summary(model)
plot(model)
```
The resulting model has the predictors

(Intercept)       
chol                
hdl                 
age                   
gendermale           
waist                
bp.1s                
BMI 
I(BMI^2)
chol:hdl         
chol:BMI         
age:bp.1s          
age:gendermale
waist:bp.1s 

With the new model, we achieve residuals which seem to be independent of the fitted
value (Residuals vs. Fitted plot) and the scale-location plot shows that the tendency of the residuals to get larger for larger fitted values decreased compared to the original model from assignment 1.

Unfortunately, some of the data (the ones including NAs had to be omitted for the model reduction). 

TODO: add log() trafo and sqrt() trafo to analysis and state that the trafos are not suitable

## Exercise 2

With the diabetes data set, perform a different analysis strategy that is based on categorizing glyhb at 2 different values, and then performing two logistic regression models with the dichotomized glyhb values.

First, discuss with your partners what could be suitable values at which glyhb could be categorized for analysis. In your considerations, take into account statistical properties of the cutpoints (enough data in all categories) and clinical relevance of the cutpoints! Let's assume your chosen cutpoints are g1 and g2.

Second, create two new binary outcome variables glyhb1 and glyhb2, where glyhb1=1 if glyhb>g1 and 0 else, or NA if glyhb is missing, and glyhb2=1 if glyhb>g2 and 0 else, or NA if glyhb is missing. Evaluate which absolute number of persons and proportions have glyhb1=1 and glyhb2=1 and glyhb1=0 and glyhb2=0 by cross-tabulating glyhb1 and glyhb2.

Third, fit two (separate) logistic regression models with glyhb1 and glyhb2 as outcome variables, using the same predictors as in Assignment 01-02. Check for nonlinear functional forms of continuous predictors. Describe the estimated associations of the predictors with the outcome variable in the model and its performance appropriately.

Report the methods and the results as you would do for a medical paper. 

Compare with your linear regression model of Exercise 1 of this assignment. How do the estimated associations of the predictors with the outcome variables compare? Are the results consistent to each other? What else can be learned here?


Discussion of suitable thresholds:
Commonly used threshold values for the glyhb value for healthy subjects are glyhb < 5.7%.
Values between 5.7% and 6.4% are the range for prediabetes, and values above 6.5% are 
are assumed to indicate diabetes. Therefore, we use g1=5.7% and g2=6.5% (see https://www.sciencedirect.com/topics/biochemistry-genetics-and-molecular-biology/glycated-hemoglobin). To check the
validity of this choice with respect to the statistical properties in our dataset, 
we display the absolute quantities of each category.

```{r}
g1 <- 5.7
g2 <- 6.5

diabetes$glyhb1 <- FALSE
diabetes$glyhb1[diabetes$glyhb > g1] <- TRUE
diabetes$glyhb1[is.na(diabetes$glyhb)] <- NA

diabetes$glyhb2 <- FALSE
diabetes$glyhb2[diabetes$glyhb > g2] <- TRUE
diabetes$glyhb2[is.na(diabetes$glyhb)] <- NA


table(diabetes$glyhb1)
```

```{r}
table(diabetes$glyhb2)
```

There seem to be sufficient instances with each value for both glyhb1 and glyhb2. Next, we show the cross-table for glyhb1 glyhb2
```{r}
table(glyhb1=diabetes$glyhb1, glyhb2=diabetes$glyhb2)
```
It is expected behavior that there are no instances with glyhb2==TRUE and 
glyhb1==FALSE, because g2>g1.
Now, we fit two logistic regression models with the same predictors as in assignment 1 to the data

```{r}
model1 <- glm(glyhb1 ~ chol + hdl + age + gender + waist + bp.1s + BMI,family=binomial(link='logit'),data=diabetes)

summary(model1)
plot(model1)
```

```{r}
model2 <- glm(glyhb2 ~ chol + hdl + age + gender + waist + bp.1s + BMI,family=binomial(link='logit'),data=diabetes)

summary(model2)
plot(model2)
``` 
The QQ-plot and scale-location plot for the second model (model2) do not look ideal. Therefore, we check for nonlinear functional forms of continuous predictors by using a 
model with interaction and quadratic terms for all predictors except for gender (assuming
the integer variables such as age are considered as continuous). Utilizing an AIC-based model reduction of a fully-quadratic model we determined the following model, which seems to perform better w.r.t to the
appearance of the QQ-plot.
```{r}
model2 <- glm(glyhb1 ~ I(bp.1s^2) + I(BMI^2) + I(age^2) + age:hdl + bp.1s:BMI + waist:BMI + waist:bp.1s + hdl:bp.1s + hdl:waist + chol:hdl + chol:bp.1s , data=diabetes,family=binomial(link='logit'))
plot(model2)
```
