---
title: "Assignment Classes 01-02"
author: "GH"
date: "2024-10-22"
output:
  word_document:
    reference_docx: MB2-Rstudio-template.docx
    toc: true
  html_document:
    df_print: paged
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions


In this exercise you will create your *first own markdown report*. You will mostly perform analyses that we have already seen in the RStudio class, but with a different data set. This exercise should help you to setup your RStudio environment properly.

Generate an R markdown report in which you perform the following analyses. Use 'MB2-Rstudio-template.docx' as template. (See the header above; it will automatically use this template, but you must have downloaded it into your R project folder.)

- Please fill in text whenever you see XXX in this file. 

- Please fill in R code whenever you see an (empty) code chunk field beginning with:

```{r}
# your R code:
```


*Finally, knit the markdown document and submit the knitted Word file!*

## Authors of this assignment

Please state name(s) and Matrikelnummer of all (up to 3) students who contributed to this assignment!

1. Michael Adam, 01507402
2. Aaron Paul Osburg, 12010313

##### page break

## Exercise 1

Find medical papers of your choice (enter the DOI) in which data analysis is used to ...

***

* ... answer a descriptive research question. 

DOI: https://doi.org/10.1038/s41591-024-03117-0

Title of the paper: Disparities in air pollution attributable mortality in the US population by race/ethnicity and sociodemographic factors

Describe the research question and give an explanation why you think it is a *descriptive* question:
This paper estimates the degree to which premature mortality in the USA
by race/ethnicity, education, rurality and social vulnerability can be
attributed to differences in exposure and susceptibility to published
particulate matter (PM2.5) air pollution estimates. This research
question is descriptive because it describes the association of
dependent variable (mortality by race, education,...) with independent
variable(s) (air pollution estimates).


***

* ... answer a predictive research question. 

DOI: https://doi.org/10.1177/2055217319885983

Title of the paper: Machine learning in secondary progressive multiple sclerosis: an improved predictive model for short-term disability progression

Describe the research question and give an explanation why you think it is a *predictive* question:
This study evaluates individual and ensemble model performance built
using decision tree (DT)-based algorithms compared to logistic
regression (LR) and support vector machines (SVMs) for predicting
secondary progressive multiple sclerosis (SPMS) disability progression.
Each of these models addresses a predictive research question because
they are designed to provide an accurate prediction of an outcome (SPMS
disability progression).


***

* ... answer a causal research question. 

DOI:  https://doi.org/10.1038/s41598-024-77131-0

Title of the paper: Causal relationships between allergic and autoimmune diseases with chronic rhinosinusitis

Describe the research question and give an explanation why you think it is a *causal* question:
This study investigates the causal relationships between allergic and
autoimmune diseases (AR, asthma, AD, and psoriasis) and Chronic
rhinosinusitis (CRS). It is a causal research question because it
explains how certain variables, in this case genetic variants associated
with autoimmune/allergic diseases, (causally) affect an outcome, in this
case being diagnosed with CRS.

##### page break

## Exercise 2

Consider that you plan to perform a regression analysis to investigate the association of an outcome variable Y with 4 independent variables X1, X2, X3, X4. What properties of the data should be checked during an "initial data analysis"?

1. The univariate distribution of the dependent and independent variables
2. The association between the independent variables
3. Missing values and patterns of missing values


##### page break

## Exercise 3






### First, download the data set

Just execute the R chunk below.

```{r}
url_diab <- "https://hbiostat.org/data/repo/diabetes.csv"
diabetes <- read.csv(file=url_diab)
head(diabetes)
```


The goal of the analysis is to estimate a descriptive model to describe the variation of `glyhb` as a linear function of `hdl`, `chol`, `age`, `gender`, `waist` circumference, `bp.1s` (systolic blood pressure)  and body-mass-index (weight in kg/height in m squared; $kg/m^2$).

### Basic description of data set

1. What is the number of observations and what is the number of variables in that data set?

```{r}
n_observations <- nrow(diabetes)
n_variables <- ncol(diabetes)
```

Fill in: 

- Number of observations: `r n_observations`
- Number of variables: `r n_variables`

### Generate new variable BMI

2. Generate a new variable `BMI` in the data set by computing it from `weight` and `height`. Note that `height` is given in inches and `weight` in lb. The correct formula is given by $BMI = weight_{lb} / height_{inch}^2 \cdot 704.5$.

```{r}
diabetes$BMI <-  diabetes$weight / diabetes$height^2 * 704.5  
```

### Missing values

3. Investigate the number of missing values per variable. Compute the number of complete observations with respect to the variables `chol`, `hdl`, `age`, `gender`, `bp1.s`, `BMI`, and `glyhb`.  

```{r}
n_missing_entries <- sapply(diabetes, function(x) sum(is.na(x)))
n_complete_obsersavtions <- n_observations - n_missing_entries
independent_variables = c("chol", "hdl", "age", "gender", "bp.1s", "BMI", "glyhb")
print(n_complete_obsersavtions[independent_variables])
```

4. Generate histograms for all independent variable of the above-described descriptive model. Generate a table with means, standard deviations and number and proportion missing for all variables, using the function of the RStudio class (`mean.sd`).

```{r}
par(mfrow = c(2, 3))
with(diabetes, {
    hist(chol)
    hist(hdl)
    hist(age)
    hist(bp.1s)
    hist(BMI)
    hist(glyhb)
})
par(mfrow = c(1, 1))


independent_variable_data <- diabetes[independent_variables]
independent_variable_data$gender[independent_variable_data$gender=="female"] <- 1
independent_variable_data$gender[independent_variable_data$gender=="male"] <- 0
independent_variable_data$gender <- as.numeric(independent_variable_data$gender)

table = data.frame(
  mean = sapply(independent_variable_data, function(x) mean(x, na.rm=T)),
  sd = sapply(independent_variable_data, function(x) sd(x, na.rm=T)),
  percentage_missing = sapply(independent_variable_data, function(x) mean(100*is.na(x)))
)

print(table)
```

5. Investigate the correlation between the independent variables by means of Spearman's correlation coefficients. You can also use code similar to the lecture, where a matrix of scatterplots and correlation coefficients is produced.

```{r}
cor(independent_variable_data, use="pairwise.complete.obs", method="spearman")
pairs(formula = ~chol+hdl+age+gender+bp.1s+BMI+glyhb , data = independent_variable_data, main = "Scatterplot Matrix")


```

6. Provide an executive summary of the initial data analysis!

Your executive summary of the initial data analysis:

The univariate distributions of the continuous independent variables chol, age, BMI and bp.1s are 
approximately normally distributed according to visual inspection of the histograms in 4.
The univariate distributions of hdl and glyhb look skewed.

Most variables have no or only weak correlations. However, stronger correlation was found between
age and bp.s1/glyhb with Spearman correlation coefficients of 0.45/0.43 (see 5.).

There are no missing values for age and gender and only very little missing values for the remaining independent variables
chol, hdl, bp1.s and BMI as well as for the dependent variable glyhb (see 3.). 

7. Estimate the linear regression model `glyhb ~ chol + hdl + age + gender + waist + bp.1s + BMI`. Provide a summary of the model. Investigate the residuals of the model. What are possible violations of this model's assumptions?

```{r}
model <- lm(formula = glyhb ~ chol + hdl + age + gender + waist + bp.1s + BMI, data=diabetes)
summary(model)
plot(model)
```
```{r}
print(mean(residuals(model)))
```

Your conclusions about possible violations of assumptions of the model:

The QQ-Plot shows that the residuals are not normally distributed.
The scale-location plot shows that the residuals tend to get larger for larger fitted values,
therefore it seems like the homoscedacity assumption is mildly violated.
Thus, not all assumptions of the linear regression model (as defined on slide
35, lecture 1) are met.

